\Title

### 2023-07-07 3.FINAL_VERSION/main.tex
Final, accepted version sent to the journal

% TODO: remove onecolumn and draftcls formatting
\documentclass[journal,twoside,web, twocolumn,draftcls]{ieeecolor}

\usepackage{generic}
\usepackage{hyperref}
\hypersetup{
    colorlinks,
    citecolor=red,
    filecolor=black,
    linkcolor=blue,
    urlcolor=black
  }
% /========== Preamble ==========/%
% Packages required by ieeecolor
\usepackage{amsmath}
%\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{textcomp}
\usepackage{cite}
\usepackage{xspace}
\usepackage{enumerate}
\usepackage{xcolor}
%---------- More packages ----------%
% NOTE: cannot use amsthm for some reason
\usepackage{mathtools}
\usepackage{bm}
\usepackage{booktabs}

@@

\definecolor{MM-blue}{HTML}{2883E6}
\newenvironment{MM}[1]{{\color{MM-blue}#1}}{}

@@

%/========== Main Document ==========/%
\begin{document}
\title{\Title}
\author{Adan Moran-MacDonald, \IEEEmembership{Member, IEEE}, Manfredi Maggiore,
\IEEEmembership{Senior Member, IEEE}*, and Xingbo Wang
\thanks{Manuscript submitted for review on \today.}
\thanks{A. Moran-MacDonald (e-mail: adan.moran@mail.utoronto.ca),
M. Maggiore (e-mail: maggiore@control.utoronto.ca), and
X. Wang (e-mail: xingbo.wang@utoronto.ca) are with the Department of
Electrical and Computer Engineering, University of Toronto, ON, Canada.}
} %/author

@@

Virtual nonholonomic constraints (\vnhcs) have been used for human-robot interaction
\cite{vnhc_human_robot_cooperation,psd_based_vnhc_redundant_manipulator,haptic_vnhc},
error-reduction on time-delayed systems \cite{vnhc_time_delay_teleop},
and they have shown marked improvements to the field of bipedal locomotion 
\cite{nhvc_dynamic_walking,
hybrid_zero_dynamics_bipedal_nhvcs,output_nhvc_bipedal_control}.
Indeed, they produce more robust walking motion in biped robots than
other virtual constraints which do not depend on velocity
\cite{nhvc_incline_walking}.
In particular, \vnhcs may be capable of injecting and
dissipating energy from a system in a robust manner, all while producing
realistic biological motion. \MM{The recent paper~\cite{SimStraBloCol23} initiates a systematic geometric investigation of \vnhcs that are affine functions of the configuration velocities.}

\MM{\textbf{Contributions of this article:}}
In this article, we design a \vnhc which provably
injects \MM{or dissipates} energy into the acrobot \MM{depending on the sign of a design parameter. The proposed constraint produces a} human-like giant motion.
\MM{The main result of the paper is presented in Theorems~\ref{thm:acrobot-oscillations} and~\ref{thm:acrobot-rotations}.   Theorem~\ref{thm:acrobot-oscillations} shows that once the virtual constraint has been enforced, for any nonzero ``low-energy'' initial condition, the proposed controller injects energy in the acrobot making it exit, in finite time, any compact subset of a certain region of the state space where the acrobot oscillates about its pivot point without performing full rotations. Theorem~\ref{thm:acrobot-rotations} handles the case when the acrobot is initialized in the complement of the above region, and it shows that under a mild assumption on the physical parameters, the acrobot  is guaranteed to keep increasing its energy while performing full rotations about the pivot point. By changing the sign of the design parameter, the results of Theorems~\ref{thm:acrobot-oscillations} and~\ref{thm:acrobot-rotations} hold for energy dissipation. By combining these energy injection/damping properties, we then propose an elementary switching logic to attain approximate energy regulation. The results of this paper are demonstrated both in simulation and experimentally. The experimental results, in particular,} demonstrate the robustness of \MM{the controller against} model uncertainty, sensor noise, and a variety of external
disturbances.

\MM{\textbf{Related literature:} While the focus of this paper is on energy injection and dissipation, a closely related concept is that of orbital stabilization to a particular energy level set. The  problem of orbital stabilization has been investigated extensively and we mention four main approaches.

@@

 There are similarities between the virtual constraint approach and the I\&I technique for orbital stabilization  proposed in~\cite{OrtYiRomAst20} in that both approaches seeks a controlled invariant submanifold on which the closed-loop system exhibit desired dynamics. But while in~\cite{OrtYiRomAst20} one wishes to obtain oscillator dynamics on the submanifold, in this paper we wish to obtain expansive or contractive dynamics (depending on the sign of a design parameter) which rules out oscillator dynamics.

In this paper, like in~\cite{YiOrtWuZha20}, we design a controller for energy pumping/damping, but the energy pumping/dumping mechanism proposed here does not arise from any form of energy shaping. In particular, it does not arise from an assignment of the damping matrix of the closed-loop system.

Finally, while in~\cite{FanLoz02} the emphasis is typically the stabilization of a specific energy level set corresponding to the homoclinic orbit of an unstable equilibrium, the technique used in this paper to inject energy does not involve the stabilization of any homoclinic orbit, and as such it does not limit the energy level that can be attained.
  }

  \MM{\textbf{Paper organization:}
In Section~\ref{sec:problem-formulation} we review the acrobot model and formulate the problem investigated in this paper. Section~\ref{sec:vnhc} presents a class of simply actuated Hamiltonian systems and preliminary notions on \vnhcs. In particular, Theorem~\ref{thm:vnhc-regularity} in this section characterizes the constrained dynamics arising from a class of \vnhcs. This result is then compared with existing literature on the subject. Section~\ref{sec:acrobot} presents the proposed \vnhc for the acrobot and the two main results of the paper, Theorems~\ref{thm:acrobot-oscillations} and~\ref{thm:acrobot-rotations} mentioned earlier. Sections~\ref{sec:simulations} and~\ref{sec:experiments} present simulations and experimental results. Section~\ref{sec:proof} presents the proof of Theorems~\ref{thm:acrobot-oscillations} and~\ref{thm:acrobot-rotations}, and Section~\ref{sec:conclusion} presents concluding remarks.
}

@@

%/========== Problem Formulation ==========/%
\section{Problem Formulation}\label{sec:problem-formulation}
We will use the simplified acrobot model in Figure
\ref{fig:acrobot}, where we assume the torso and leg rods are of
equal length \(l\) with equal point masses \(m\) at the tips.
The acrobot's configuration is described in generalized coordinates
\MM{\(q=(q_u,q_a)\)} on the configuration manifold 
\(\mathcal{Q} = \Sone \times \Sone\), where \(q_u\) is unactuated and 
\(q_a\) is actuated.
We ignore dissipative forces in this model.

\MM{In what follows, we denote by \(c_u \eqdef \cos(q_u)\), \(c_a \eqdef \cos(q_a)\), and 
\(c_{ua} \eqdef \cos(q_u + q_a)\); likewise, \(s_u \eqdef \sin(q_u)\), 
\(s_a \eqdef \sin(q_a)\), and \(s_{ua} \eqdef \sin(q_u + q_a)\).
}

The acrobot has inertia matrix 
\(M\), potential function \(V\) (with respect to
the horizontal bar), and input matrix \(B\) given as follows:
\begin{align}\label{eqn:acrobot-inertia}
    M(q) &= \begin{bmatrix}
        ml^2\left(3+2 c_a\right) & 
        ml^2\left(1+ c_a \right) \\
        ml^2\left(1+ c_a\right) &
        ml^2
    \end{bmatrix} 
    , \\
    \label{eqn:acrobot-potential}
    V(q) &= -mgl\left(2 c_u + c_{ua}\right)
    , \\
    \label{eqn:acrobot-B}
    B &= [0;1]
    .
\end{align}
For reasons that will become clear later in this article, we
use Hamiltonian mechanics to derive the dynamics of the acrobot. 
\MM{To this end, recalling the partitioning  \(q= (q_u,q_a)\), we define the conjugate of momenta, \(p = (p_u,p_a) = M(q)\dot{q}\) and the Hamiltonian function
\begin{equation}\label{eqn:acrobot-hamiltonian}
    \mathcal{H}(q,p) = \frac{1}{2}p\tpose \Minv(q) p -
   mgl\left(2 c_u + c_{ua}\right).
  \end{equation}
The dynamics of the
acrobot with Hamiltonian function~\eqref{eqn:acrobot-hamiltonian} are
\begin{equation}
  \label{eq:acrobot:model}
  \begin{aligned}
&    \dot{q} = \Minv(q) p,\\
&    \dot{p}_u = -mgl\left(2s_u + s_{ua}\right),\\
&     \dot{p}_a =-\frac{1}{2}p\tpose \nabla_{q_a}\Minv(q) p - m g l s_{ua} + \tau,
  \end{aligned}
\end{equation}
where \(\tau\), the control input, is the hip joint torque affecting only the dynamics of
\(p_a\).}

@@

%/========== \vnhc ==========/%
\section{Preliminaries on \vnhcs}\label{sec:vnhc}
Before embarking on the design problem, we must summarize the relevant
theory of \vnhcs for a class of mechanical systems we
call ``simply actuated Hamiltonian systems". 
\MM{Although the results in Section \ref{sec:vnhc-vnhc} do not appear in the literature in the same form, we will show that} they are a special case of results that appeared in
\cite{hybrid_zero_dynamics_bipedal_nhvcs}.

\subsection{Simply Actuated Hamiltonian Systems} \label{sec:vnhc-sah}
\MM{Consider} a mechanical system modelled with generalized coordinates 
\(q = (q_1, \ldots, q_n)\) on a configuration manifold
\(\mathcal{Q} = \Rt{T_1} \times \cdots \times \Rt{T_n}\), where
\(T_i = 2\pi\) if \(q_i\) is an angle and \(T_i = \infty\) if \(q_i\) is a
displacement. The corresponding generalized velocities are 
\(\dot{q} = (\dot{q}_1,\ldots,\dot{q}_n) \in \R^n\).

Suppose this system has Lagrangian
\(\mathcal{L}(q,\dot{q}) = 1/2~\dot{q}^T D(q) \dot{q} - P(q)\),
where the potential energy 
\(P : \mathcal{Q} \rightarrow \mathbb{R}\) 
is smooth, and the inertia matrix 
\(D : \mathcal{Q} \rightarrow \mathbb{R}^{n \times n}\)
is smooth and positive definite for all \(q \in \mathcal{Q}\).
The \textit{conjugate momentum} to \(q\) is the vector
\(p \eqdef \partial\mathcal{L}/\partial\dot{q} = D(q) \dot{q} \in \R^n\).
As per \cite{landau_mechanics}, 
the \textit{Hamiltonian} of the system in \((q,p)\) coordinates
is
\begin{equation}\label{eqn:hamiltonian}
    \mathcal{H}(q,p) = \frac{1}{2} p\tpose D\inv(q) p + P(q)
    ,
\end{equation}
with dynamics
\begin{equation}\label{eqn:hamiltonian-eom-general}
    \begin{cases}
        \dot{q} = \nabla_p\mathcal{H} 
        , \\
        \dot{p} = -\nabla_q\mathcal{H} + \MM{B} \, \tau
        ,
    \end{cases}
\end{equation}
where \(\tau \in \R^k\) is a vector of generalized input forces and the input
matrix \MM{\(B \in \R^{n \times k}\) is assumed to be constant and of full rank \(k\).}
If \(k < n\), we say the system is \textit{underactuated} with degree of
underactuation \((n-k)\).

Using the matrix Kronecker product, it is easy to show that
\eqref{eqn:hamiltonian-eom-general} expands to
\begin{equation*}\label{eqn:hamiltonian-full-dynamics}
     \begin{cases}
        \dot{q} = D\inv(q)p 
        , \\
        \dot{p} = -\frac{1}{2} (\Id{n} \otimes p\tpose) \nabla_q D\inv(q) p
        - \nabla_q P(q) + \MM{B} \,\tau
        . \\
    \end{cases}
\end{equation*}
%\begin{align*}
%    \dot{q} &= D\inv(q)p 
%    ,\\
%    \dot{p} &= -\frac{1}{2} (\Id{n} \otimes p\tpose) \nabla_q D\inv(q) p
%        - \nabla_q P(q) + B(q) \tau
%    .
%\end{align*}

Because \(\tau\) is transformed by \(\MM{B}\), it is not obvious how any
particular input force \(\tau_i\) affects the system.
% As a first step in addressing this problem, we make the following assumptions.

% \begin{assm}\label{assm:B-const}
%     The input matrix \(B(q) \equiv B \in \R^{n\times k}\) is constant
%     and has full rank \(k < n\).
% \end{assm}
%
When \(\mathcal{Q} = \R^n\), \MM{one may} define a
canonical coordinate transformation of \eqref{eqn:hamiltonian} 
which decouples the input forces. To define this transformation we will make use of the following lemma\footnote{\MM{Lemma~\ref{lemma:B-orthogonal} and Theorem~\ref{thm:simply-actuated} are obvious, but we include their proofs for completeness.}}.

@@

\begin{thm}\label{thm:simply-actuated}
\MM{For} the Hamiltonian system \eqref{eqn:hamiltonian} with configuration
    manifold \(\mathcal{Q} = \R^n\)\MM{,
    the} coordinate transformation
    \((\tilde q,\tilde p) = \left(\mathbf{B}\tpose)\inv q,\mathbf{B}p\right)\)
    is a canonical transformation. \MM{ the} resulting \MM{Hamiltonian function is 
    \begin{equation}
    \label{eqn:simple-hamiltonian}
        \mathcal{H}(\tilde{q},\tilde{p}) = 
        \frac{1}{2} \tilde{p}\tpose \Minv(\tilde{q}) \tilde{p} + V(\tilde{q}),
      \end{equation}
      and the associated Hamiltonian dynamics are
      \begin{equation}
        \label{eq:Hamiltonian_system:simply_actuated}
          \begin{aligned}
            &           \dot{\tilde{q}} = \Minv(\tilde{q})\tilde{p}, \\
            &           \dot{\tilde{p}} = -\frac{1}{2} (\Id{n} \otimes \tilde{p}\tpose)
              \nabla_{\tilde{q}} \Minv(\tilde{q}) \tilde{p} \\
            &          \phantom{---} - \nabla_{\tilde{q}} V(\tilde{q}) + \simpleB \tau,
          \end{aligned}
  \end{equation}}
  where 
    \(\Minv(\tilde{q}) \eqdef 
    (\mathbf{B}\tpose)\inv D^{-1}(\mathbf{B}\tpose \tilde{q})\mathbf{B}\inv\)
    and
    \(V(\tilde{q}) \eqdef P(\mathbf{B}\tpose \tilde{q})\).
\end{thm}
\begin{proof}
    Since \(\mathbf{B}\) is constant, this transformation satisfies
    \(\partial\tilde{q}_i/\partial p_j = \partial\tilde{p}_i/\partial q_j = 0\) for all 
    \(i,j \in \{1,\ldots,n\}\).
    This implies the Poisson brackets \([\tilde{q}_i, \tilde{q}_j]\)
    and \([\tilde{p}_i,\tilde{p}_j]\) are both zero.
    Then, one can show that
    \([\tilde{p}_i, \tilde{q}_j] = (\mathbf{B}\tpose)\inv_i (\mathbf{B}\tpose)_j
        = \delta_i^j\).
    By (45.10) in \cite{landau_mechanics}, this is a canonical transformation
    and the new Hamiltonian is
    \(\mathcal{H}(\mathbf{B}\tpose \tilde{q}, \mathbf{B}\inv \tilde{p})\).
    Finally, since \(\dot{\tilde{p}} = \mathbf{B} \dot{p}\), the input
    matrix for the system in \((\tilde{q},\tilde{p})\) coordinates is
    \(\mathbf{B}B = [\Zmat{(n-k)\times k}; \Id{k}]\), which proves the theorem.
\end{proof}

We call the \((\tilde{q},\tilde{p})\) coordinates
\textit{simply actuated coordinates}, and we call any Hamiltonian system
\MM{of the form~\eqref{eq:Hamiltonian_system:simply_actuated}} a 
\textit{simply actuated Hamiltonian system}.
The first \((n-k)\) configuration variables in \(\tilde{q}\), labelled \(q_u\),
are the \textit{unactuated coordinates}; 
the remaining \(k\) configuration variables, labelled \(q_a\), are the
\textit{actuated coordinates}.
The corresponding \((p_u, p_a)\) in \(\tilde{p}\) are the \textit{unactuated}
and \textit{actuated momenta}, respectively.

@@

Griffin and Grizzle \cite{nhvc_dynamic_walking} were the first to define
relative degree two nonholonomic constraints which can be enforced through state
feedback. Horn \etal extended their results in
\cite{hybrid_zero_dynamics_bipedal_nhvcs} to derive the constrained dynamics for
a certain class of mechanical systems. 
Hamed and Ames \cite{nonholonomic_hybrid_zero_dynamics} then found an explicit
representation of these constrained dynamics for \vnhcs with a specific form.
All these researchers made use of the
unactuated conjugate momentum, but they developed their results in the
Lagrangian framework. In particular, they focused on Lagrangian systems with
degree of underactuation one. \MM{The recent paper~\cite{SimStraBloCol23} has initiated a systematic geometric investigation of \vnhcs that are affine functions of the configuration velocities. The ones used in this paper are nonlinear functions of the configuration velocities.}

@@

\begin{defn}\label{defn:vnhc}
    A \textit{virtual nonholonomic constraint} (\vnhc) \textit{of order \(k\)}
    is a relation \(h(q,p) = 0\) where 
    \(h : \mathcal{Q}\times\R^n \rightarrow \R^k\) is \(C^2\),
    \(\rank{\left[ dh_q,\, dh_p \right]} = k\) for all 
    \((q,p) \in h\inv(0)\), and there exists a feedback
    controller \(\tau(q,p)\) rendering the \textit{constraint manifold}
    \(\Gamma\) invariant, where
    \begin{equation}
        \Gamma = \left\{(q,p) \mid h(q,p) = 0, dh_q \dot{q} + dh_p \dot{p} = 0\right\}
        .
    \end{equation}
\end{defn}

@@

\begin{defn}
    A \vnhc \(h(q,p) = 0\) of order \(k\) is \textit{regular} if \MM{system~\eqref{eq:Hamiltonian_system:simply_actuated}} with output 
    \(e = h(q,p)\)  \MM{has vector} relative degree \(\{2,2.\ldots,2\}\) everywhere on the
    constraint manifold \(\Gamma\).
\end{defn}

>>>>

% TODO: remove onecolumn and draftcls formatting
\documentclass[journal,twoside,web, twocolumn]{ieeecolor}

\usepackage{generic}
\usepackage[hypertexnames=false]{hyperref}
\hypersetup{
    colorlinks,
    citecolor=red,
    filecolor=black,
    linkcolor=blue,
    urlcolor=black
  }
% /========== Preamble ==========/%
% Packages required by ieeecolor
\usepackage{amsmath}
%\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{textcomp}
\usepackage{cite}
\usepackage{xspace}
\usepackage{enumerate}
%\usepackage{xcolor}
%---------- More packages ----------%
% NOTE: cannot use amsthm for some reason
\usepackage{mathtools}
\usepackage{bm}
\usepackage{booktabs}

@@

@@

%/========== Main Document ==========/%
\begin{document}
\title{\Title}
\author{Adan Moran-MacDonald, \IEEEmembership{Member, IEEE}, Manfredi Maggiore,
\IEEEmembership{Senior Member, IEEE}, and Xingbo Wang
\thanks{A. Moran-MacDonald (e-mail: adan.moran@mail.utoronto.ca),
M. Maggiore (e-mail: maggiore@control.utoronto.ca), and
X. Wang (e-mail: xingbo.wang@utoronto.ca) are with the Department of
Electrical and Computer Engineering, University of Toronto, ON, Canada. This research was supported by the Natural Sciences and Engineering Research Council of Canada (NSERC).}
} %/author

@@

Virtual nonholonomic constraints (\vnhcs) have been used for human-robot interaction
\cite{vnhc_human_robot_cooperation,psd_based_vnhc_redundant_manipulator,haptic_vnhc},
error-reduction on time-delayed systems \cite{vnhc_time_delay_teleop},
and they have shown marked improvements to the field of bipedal locomotion 
\cite{nhvc_dynamic_walking,
hybrid_zero_dynamics_bipedal_nhvcs,output_nhvc_bipedal_control}.
Indeed, they produce more robust walking motion in biped robots than
other virtual constraints which do not depend on velocity
\cite{nhvc_incline_walking}.
In particular, \vnhcs may be capable of injecting and
dissipating energy from a system in a robust manner, all while producing
realistic biological motion. The recent paper~\cite{SimStraBloCol23} initiates a systematic geometric investigation of \vnhcs that are affine functions of the configuration velocities.

\textbf{Contributions of this article:}
In this article, we design a \vnhc which provably
injects or dissipates energy into the acrobot depending on the sign of a design parameter. The proposed constraint produces a human-like giant motion.
The main result of the paper is presented in Theorems~\ref{thm:acrobot-oscillations} and~\ref{thm:acrobot-rotations}.   Theorem~\ref{thm:acrobot-oscillations} shows that once the virtual constraint has been enforced, for any nonzero ``low-energy'' initial condition, the proposed controller injects energy in the acrobot making it exit, in finite time, any compact subset of a certain region of the state space where the acrobot oscillates about its pivot point without performing full rotations. Theorem~\ref{thm:acrobot-rotations} handles the case when the acrobot is initialized in the complement of the above region, and it shows that under a mild assumption on the physical parameters, the acrobot  is guaranteed to keep increasing its energy while performing full rotations about the pivot point. By changing the sign of the design parameter, the results of Theorems~\ref{thm:acrobot-oscillations} and~\ref{thm:acrobot-rotations} hold for energy dissipation. By combining these energy injection/damping properties, we then propose an elementary switching logic to attain approximate energy regulation. The results of this paper are demonstrated both in simulation and experimentally. The experimental results, in particular, demonstrate the robustness of the controller against model uncertainty, sensor noise, and a variety of external
disturbances.

\textbf{Related literature:} While the focus of this paper is on energy injection and dissipation, a closely related concept is that of orbital stabilization to a particular energy level set. The  problem of orbital stabilization has been investigated extensively and we mention four main approaches.

@@

 There are similarities between the virtual constraint approach and the I\&I technique for orbital stabilization  proposed in~\cite{OrtYiRomAst20} in that both approaches seek a controlled invariant submanifold on which the closed-loop system exhibits desired dynamics. But while in~\cite{OrtYiRomAst20} one wishes to obtain oscillator dynamics on the submanifold, in this paper we wish to obtain expansive or contractive dynamics (depending on the sign of a design parameter) which rules out oscillator dynamics.

In this paper, like in~\cite{YiOrtWuZha20}, we design a controller for energy pumping/damping, but the energy pumping/damping mechanism proposed here does not arise from any form of energy shaping. In particular, it does not arise from an assignment of the damping matrix of the closed-loop system.

Finally, while in~\cite{FanLoz02} the emphasis is typically the stabilization of a specific energy level set corresponding to the homoclinic orbit of an unstable equilibrium, the technique used in this paper to inject energy does not involve the stabilization of any homoclinic orbit, and as such it does not limit the energy level that can be attained.

\textbf{Paper organization:}
In Section~\ref{sec:problem-formulation} we review the acrobot model and formulate the problem investigated in this paper. Section~\ref{sec:vnhc} presents a class of simply actuated Hamiltonian systems and preliminary notions on \vnhcs. In particular, Theorem~\ref{thm:vnhc-regularity} in this section characterizes the constrained dynamics arising from a class of \vnhcs. This result is then compared with existing literature on the subject. Section~\ref{sec:acrobot} presents the proposed \vnhc for the acrobot and the two main results of the paper, Theorems~\ref{thm:acrobot-oscillations} and~\ref{thm:acrobot-rotations} mentioned earlier. Sections~\ref{sec:simulations} and~\ref{sec:experiments} present simulations and experimental results. Section~\ref{sec:proof} presents the proof of Theorems~\ref{thm:acrobot-oscillations} and~\ref{thm:acrobot-rotations}, and Section~\ref{sec:conclusion} presents concluding remarks.

@@

%/========== Problem Formulation ==========/%
\section{Problem Formulation}\label{sec:problem-formulation}
We will use the simplified acrobot model in Figure \ref{fig:acrobot}, where we assume the torso and leg rods are of equal length \(l\) with equal point masses \(m\) at the tips. The acrobot's configuration is described in generalized coordinates \(q=(q_u,q_a)\) on the configuration manifold \(\mathcal{Q} = \Sone \times \Sone\), where \(q_u\) is unactuated and  \(q_a\) is actuated. We ignore dissipative forces in this model.

In what follows, we denote by \(c_u \eqdef \cos(q_u)\), \(c_a \eqdef \cos(q_a)\), and \(c_{ua} \eqdef \cos(q_u + q_a)\); likewise, \(s_u \eqdef \sin(q_u)\), \(s_a \eqdef \sin(q_a)\), and \(s_{ua} \eqdef \sin(q_u + q_a)\).

The acrobot has inertia matrix \(M\), potential function \(V\) (with respect to the horizontal bar), and input matrix \(B\) given as follows:
\begin{align}\label{eqn:acrobot-inertia}
    M(q) &= \begin{bmatrix}
        ml^2\left(3+2 c_a\right) & 
        ml^2\left(1+ c_a \right) \\
        ml^2\left(1+ c_a\right) &
        ml^2
    \end{bmatrix} 
    , \\
    \label{eqn:acrobot-potential}
    V(q) &= -mgl\left(2 c_u + c_{ua}\right)
    , \\
    \label{eqn:acrobot-B}
    B &= [0;1]
    .
\end{align}
For reasons that will become clear later in this article, we
use Hamiltonian mechanics to derive the dynamics of the acrobot. 
To this end, recalling the partitioning  \(q= (q_u,q_a)\), we define the conjugate of momenta, \(p = (p_u,p_a) = M(q)\dot{q}\) and the Hamiltonian function
\begin{equation}\label{eqn:acrobot-hamiltonian}
    \mathcal{H}(q,p) = \frac{1}{2}p\tpose \Minv(q) p -
   mgl\left(2 c_u + c_{ua}\right).
  \end{equation}
The dynamics of the
acrobot with Hamiltonian function~\eqref{eqn:acrobot-hamiltonian} are
\begin{equation}
  \label{eq:acrobot:model}
  \begin{aligned}
&    \dot{q} = \Minv(q) p,\\
&    \dot{p}_u = -mgl\left(2s_u + s_{ua}\right),\\
&     \dot{p}_a =-\frac{1}{2}p\tpose \nabla_{q_a}\Minv(q) p - m g l s_{ua} + \tau,
  \end{aligned}
\end{equation}
where \(\tau\), the control input, is the hip joint torque affecting only the dynamics of
\(p_a\).

@@

%/========== \vnhc ==========/%
\section{Preliminaries on \vnhcs}\label{sec:vnhc}
Before embarking on the design problem, we must summarize the relevant theory of \vnhcs for a class of mechanical systems we call ``simply actuated Hamiltonian systems". Although the results in Section \ref{sec:vnhc-vnhc} do not appear in the literature in the same form, we will show that they are a special case of results that appeared in \cite{hybrid_zero_dynamics_bipedal_nhvcs}.

\subsection{Simply Actuated Hamiltonian Systems} \label{sec:vnhc-sah}
Consider a mechanical system modelled with generalized coordinates  \(q = (q_1, \ldots, q_n)\) on a configuration manifold \(\mathcal{Q} = \Rt{T_1} \times \cdots \times \Rt{T_n}\), where \(T_i = 2\pi\) if \(q_i\) is an angle and \(T_i = \infty\) if \(q_i\) is a displacement. The corresponding generalized velocities are \(\dot{q} = (\dot{q}_1,\ldots,\dot{q}_n) \in \R^n\).

Suppose this system has Lagrangian \(\mathcal{L}(q,\dot{q}) = 1/2~\dot{q}^T D(q) \dot{q} - P(q)\), where the potential energy  \(P : \mathcal{Q} \rightarrow \mathbb{R}\)  is smooth, and the inertia matrix  \(D : \mathcal{Q} \rightarrow \mathbb{R}^{n \times n}\) is smooth and positive definite for all \(q \in \mathcal{Q}\). The \textit{conjugate momentum} to \(q\) is the vector \(p \eqdef \partial\mathcal{L}/\partial\dot{q} = D(q) \dot{q} \in \R^n\). As per \cite{landau_mechanics},  the \textit{Hamiltonian} of the system in \((q,p)\) coordinates is
\begin{equation}\label{eqn:hamiltonian}
    \mathcal{H}(q,p) = \frac{1}{2} p\tpose D\inv(q) p + P(q)
    ,
\end{equation}
with dynamics
\begin{equation}\label{eqn:hamiltonian-eom-general}
    \begin{cases}
        \dot{q} = \nabla_p\mathcal{H} 
        , \\
        \dot{p} = -\nabla_q\mathcal{H} + B \, \tau
        ,
    \end{cases}
\end{equation}
where \(\tau \in \R^k\) is a vector of generalized input forces and the input matrix \(B \in \R^{n \times k}\) is assumed to be constant and of full rank \(k\). If \(k < n\), we say the system is \textit{underactuated} with degree of underactuation \((n-k)\).

Using the matrix Kronecker product, it is easy to show that
\eqref{eqn:hamiltonian-eom-general} expands to
\begin{equation*}\label{eqn:hamiltonian-full-dynamics}
     \begin{cases}
        \dot{q} = D\inv(q)p 
        , \\
        \dot{p} = -\frac{1}{2} (\Id{n} \otimes p\tpose) \nabla_q D\inv(q) p
        - \nabla_q P(q) + B \,\tau
        . \\
    \end{cases}
\end{equation*}
%\begin{align*}
%    \dot{q} &= D\inv(q)p 
%    ,\\
%    \dot{p} &= -\frac{1}{2} (\Id{n} \otimes p\tpose) \nabla_q D\inv(q) p
%        - \nabla_q P(q) + B(q) \tau
%    .
%\end{align*}

Because \(\tau\) is transformed by \(B\), it is not obvious how any particular input force \(\tau_i\) affects the system.

When \(\mathcal{Q} = \R^n\), one may define a canonical coordinate transformation of \eqref{eqn:hamiltonian} which decouples the input forces. To define this transformation we will make use of the following lemma\footnote{Lemma~\ref{lemma:B-orthogonal} and Theorem~\ref{thm:simply-actuated} are obvious, but we include their proofs for completeness.}.

@@

\begin{thm}\label{thm:simply-actuated}
For the Hamiltonian system \eqref{eqn:hamiltonian} with configuration manifold \(\mathcal{Q} = \R^n\), the coordinate transformation \((\tilde q,\tilde p) = \left(\mathbf{B}\tpose)\inv q,\mathbf{B}p\right)\) is a canonical transformation. The resulting Hamiltonian function is 
    \begin{equation}
    \label{eqn:simple-hamiltonian}
        \mathcal{H}(\tilde{q},\tilde{p}) = 
        \frac{1}{2} \tilde{p}\tpose \Minv(\tilde{q}) \tilde{p} + V(\tilde{q}),
      \end{equation}
      and the associated Hamiltonian dynamics are
      \begin{equation}
        \label{eq:Hamiltonian_system:simply_actuated}
          \begin{aligned}
            &           \dot{\tilde{q}} = \Minv(\tilde{q})\tilde{p}, \\
            &           \dot{\tilde{p}} = -\frac{1}{2} (\Id{n} \otimes \tilde{p}\tpose)
              \nabla_{\tilde{q}} \Minv(\tilde{q}) \tilde{p} \\
            &          \phantom{---} - \nabla_{\tilde{q}} V(\tilde{q}) + \simpleB \tau,
          \end{aligned}
  \end{equation}
  where 
    \(\Minv(\tilde{q}) \eqdef 
    (\mathbf{B}\tpose)\inv D^{-1}(\mathbf{B}\tpose \tilde{q})\mathbf{B}\inv\)
    and
    \(V(\tilde{q}) \eqdef P(\mathbf{B}\tpose \tilde{q})\).
\end{thm}
\begin{proof}
    Since \(\mathbf{B}\) is constant, this transformation satisfies
    \(\partial\tilde{q}_i/\partial p_j = \partial\tilde{p}_i/\partial q_j = 0\) for all 
    \(i,j \in \{1,\ldots,n\}\).
    This implies the Poisson brackets \([\tilde{q}_i, \tilde{q}_j]\)
    and \([\tilde{p}_i,\tilde{p}_j]\) are both zero.
    Then, one can show that
    \([\tilde{p}_i, \tilde{q}_j] = (\mathbf{B}\tpose)\inv_i (\mathbf{B}\tpose)_j
        = \delta_i^j\).
    By (45.10) in \cite{landau_mechanics}, this is a canonical transformation
    and the new Hamiltonian is
    \(\mathcal{H}(\mathbf{B}\tpose \tilde{q}, \mathbf{B}\inv \tilde{p})\).
    Finally, since \(\dot{\tilde{p}} = \mathbf{B} \dot{p}\), the input
    matrix for the system in \((\tilde{q},\tilde{p})\) coordinates is
    \(\mathbf{B}B = [\Zmat{(n-k)\times k}; \Id{k}]\), which proves the theorem.
\end{proof}

We call the \((\tilde{q},\tilde{p})\) coordinates \textit{simply actuated coordinates}, and we call any Hamiltonian system of the form~\eqref{eq:Hamiltonian_system:simply_actuated} a  \textit{simply actuated Hamiltonian system}. The first \((n-k)\) configuration variables in \(\tilde{q}\), labelled \(q_u\), are the \textit{unactuated coordinates};  the remaining \(k\) configuration variables, labelled \(q_a\), are the \textit{actuated coordinates}. The corresponding \((p_u, p_a)\) in \(\tilde{p}\) are the \textit{unactuated} and \textit{actuated momenta}, respectively.

@@

Griffin and Grizzle \cite{nhvc_dynamic_walking} were the first to define relative degree two nonholonomic constraints which can be enforced through state feedback. Horn \etal extended their results in \cite{hybrid_zero_dynamics_bipedal_nhvcs} to derive the constrained dynamics for a certain class of mechanical systems.  Hamed and Ames \cite{nonholonomic_hybrid_zero_dynamics} then found an explicit representation of these constrained dynamics for \vnhcs with a specific form. All these researchers made use of the unactuated conjugate momentum, but they developed their results in the Lagrangian framework. In particular, they focused on Lagrangian systems with degree of underactuation one. The recent paper~\cite{SimStraBloCol23} has initiated a systematic geometric investigation of \vnhcs that are affine functions of the configuration velocities. The ones used in this paper are nonlinear functions of the configuration velocities.

@@

\begin{defn}\label{defn:vnhc}
    A \textit{virtual nonholonomic constraint} (\vnhc) \textit{of order \(k\)}
    is a relation \(h(q,p) = 0\) where 
    \(h : \mathcal{Q}\times\R^n \rightarrow \R^k\) is \(C^2\),
    \(\rank{\left[ dh_q,\, dh_p \right]} = k\) for all 
    \((q,p) \in h\inv(0)\), and there exists a feedback
    controller \(\tau(q,p)\) rendering the \textit{constraint manifold}
    \(\Gamma\) invariant, where
\begin{equation}
        \Gamma = \left\{(q,p) \mid h(q,p) = 0, dh_q \dot{q} + dh_p \dot{p} = 0\right\}.    \end{equation}
\end{defn}

@@

\begin{defn}
A \vnhc \(h(q,p) = 0\) of order \(k\) is \textit{regular} if system~\eqref{eq:Hamiltonian_system:simply_actuated} with output \(e = h(q,p)\) has vector relative degree \(\{2,2.\ldots,2\}\) everywhere on the constraint manifold \(\Gamma\).
\end{defn}

The reader is referred to~\cite{Isi95} for the definition of vector relative degree. The authors of \cite{nhvc_dynamic_walking,hybrid_zero_dynamics_bipedal_nhvcs} observed that relations which use only the unactuated conjugate momentum often have vector relative degree \(\{2,\ldots,2\}\). Indeed, we shall now prove that regular constraints cannot depend on the actuated momentum.
